---
title: "Data Notebook"
format: html
editor: visual
---

```{r}
#A - Load libraries
library(tidyverse)
library(rio)
library(janitor)
library(dplyr)

```

```{r}
#B - Import the data

CFPB <- rio::import("complaints-2025-04-29_20_45.csv")

CFPB <- clean_names(CFPB)
```

```{r}
# 1. Complaints by Month
# Convert date to proper format and extract month
CFPB <- CFPB %>%
  mutate(date_received = mdy(date_received),
         month = floor_date(date_received, "month"))

# Group by month and count complaints
complaints_by_month <- CFPB %>%
  group_by(month) %>%
  summarise(complaint_count = n()) %>%
  arrange(month)

# View the monthly complaints
print(complaints_by_month)
```

```{r}
complaints_by_state <- CFPB %>%
  group_by(state) %>%
  summarise(complaint_count = n()) %>%
  arrange(desc(complaint_count))

# View the state complaints
print(complaints_by_state)
```

```{r}
write.csv(complaints_by_state, "complaints_by_state.csv", row.names = FALSE)
write.csv(complaints_by_month, "complaints_by_month.csv", row.names = FALSE)
```

```{r}


# Group products into the specified categories
complaints_grouped <- CFPB %>%
  mutate(product_category = case_when(
    grepl("student loan", product, ignore.case = TRUE) ~ "Student loan",
    grepl("credit reporting|personal consumer report", product, ignore.case = TRUE) ~ "Credit reporting or other personal consumer reports",
    grepl("debt collection", product, ignore.case = TRUE) ~ "Debt collection",
    TRUE ~ "Other"
  ))

# Check the distribution
product_summary <- complaints_grouped %>%
  group_by(product_category) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

# View the summary
print(product_summary)

write.csv(product_summary, "complaints_by_type.csv", row.names = FALSE)

